How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?

### 2024-02-10 outputs/paper/paper.qmd
Made figure 2 and 3

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace setup ####
library(tidyverse)
library(sf)
library(readxl)
library(knitr)
library(janitor)
library(lubridate)
library(dplyr)
library(data.table)
library(RColorBrewer)
library(ggpubr)
library(scales)

@@

ghs_data <- read_excel(here::here("inputs/data/GHS_index.xlsx"))
world_map <- st_read(here::here("inputs/data/world/World_Countries__Generalized_.shp"))
```

@@

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
library(tidyverse)
library(sf)
library(readxl)
library(knitr)
library(janitor)
library(lubridate)
library(dplyr)
library(data.table)
library(RColorBrewer)
library(ggpubr)
library(scales)

@@

ghs_data <- read_excel(here::here("inputs/data/GHS_index.xlsx"))
world_map <- st_read(here::here("inputs/data/world/World_Countries__Generalized_.shp"))
life_exp <- read.csv(here::here("inputs/data/life_table.csv"), stringsAsFactors = FALSE)
```

@@

```{r fig.width=7, fig.height=3.5}
#| echo: false
#| eval: true
#| label: fig-ELE
#| fig-cap: Estimates of Life Expectancy at Birth, by Race 2006-2021
#| warning: false

# Reshape the data if necessary
life_exp_melted <- reshape(
  life_exp,
  idvar = "year_id",
  varying = list(names(life_exp)[-1]), 
  v.names = "life_exp",
  times = names(life_exp)[-1],
  new.row.names = 1:1e5,
  direction = "long"
)
life_exp_melted$race <- gsub("_", " ", life_exp_melted$time)  # Convert "_" to " " in race names
life_exp_melted <- life_exp_melted[, c("year_id", "race", "life_exp")]

life_exp_melted[life_exp_melted$race == 'All', "race"] <- 'All races and origins'

## CREATE PANEL A
est_life <- ggplot(data = life_exp_melted[!(life_exp_melted$race %like% 'AIAN|Asian'), ], aes(x = year_id, y = life_exp, color = race)) +
  geom_point(shape = 16, size = 3.5, alpha = 0.9) +  # Circular points with size 2
  geom_line(aes(group = race), alpha = 0.56, size = 1.5) +  # Add group aesthetic for lines
  scale_y_continuous(name = 'Life expectancy at birth', limits = c(65, 85)) +
  scale_x_continuous(name = '', breaks = seq(2006, 2020, by = 2), limits = c(2005.5, 2021.5)) +
  scale_color_manual(values = c('#0c4928', '#0b3669', '#4c2962', '#5a0f1c'), 
                     labels = c("All races and origins", "Hispanic", "Black", "White")) +
  # Darker color values
  theme_minimal() +
  theme(axis.title = element_text(size = 14),  # Adjusting the size of the y-axis label
        axis.text = element_text(size = 12),
        legend.position = 'bottom',  # Set legend position to bottom
        legend.justification = 'center',  # Center the legend
        legend.text = element_text(size = 10),  # Adjusting the font size for legend text
        legend.title = element_text(size = 12, face = 'bold')) +
  labs(color = "Race:")  # Set the legend title

est_life
# Save the plot to a file
ggsave("Life_Exp.png", width = 10, height = 8)
```

```{r}
#| echo: false
#| eval: true
#| warning: false
# Create a dataframe to store the changes by race
change_by_race <- data.frame(
  race = c("All races and origins", "Hispanic", "American Indians and Alaska Native", "Asian", "Black", "White"),
  change_2020_2019 = c(77 - 78.8, 77.9 - 81.9, 67.1 - 71.8, 83.6 - 85.6, 71.5 - 74.8, 77.4 - 78.8),
  change_2021_2020 = c(76.1 - 77, 77.7 - 77.9, 65.2 - 67.1, 83.5 - 83.6, 70.8 - 71.5, 76.4 - 77.4)
)

```

```{r fig.width=8, fig.height=4}
#| echo: false
#| eval: true
#| label: fig-CLE
#| fig-cap: Change in Life Expectancy at Birth from the Previous Year
#| warning: false

# Melt the dataframe
tmp <- melt(change_by_race, id.vars = "race", measure.vars = c("change_2020_2019", "change_2021_2020"), 
             variable.name = "period", value.name = "diff")

# Create PANEL B
change_life_exp <- ggplot(data = tmp, aes(x = race, y = diff, fill = period)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.71) +
  scale_y_continuous(name = "Change (years)", breaks = -5:0, limits = c(-5.5, 0)) +
  scale_fill_manual(name = "Time period", values = c('#243f29', '#186d38'), labels = c('2019-2020', '2020-2021')) + 
  scale_color_manual(name = "Time period", values = c('#243f29', '#186d38'), labels = c('2019-2020', '2020-2021')) + 
  theme_minimal() + labs(x = '') +
  geom_text(aes(label = sprintf("%.1f", diff)), position = position_dodge(width = 0.9), vjust = 1.3, size = 4) +
  theme(
    axis.title = element_text(size = 14),
    axis.text.x = element_text(size = 10, hjust = 1),
    axis.text.y = element_text(size = 8),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 12, face = 'bold')) + 
    scale_x_discrete(labels = function(x) str_wrap(x, width = 15))

change_life_exp
ggsave("Change_Life_Exp.png", width = 10, height = 8)

```
